---
title: "Thoughts on spurious correlations driving correlations between basal and acclimation responses"
date: "3/19/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Toy example of a spurious correlation between unrelated variables

Here I randomly generate values from a normal distribution to represent independent basal and acclimation variables. Next I substract the basal value from the acclimation value to get our capacity variable (similar to what was done in the Gerken paper). Finally, we examine the correlation between the capacity variable and the basal variable.
```{r}

# Simulated variables drawn from a normal dist. with identical variance (sd)
basal <- rnorm(100,0,1)
acclimation <- rnorm(100,0,1)
# Calculate capacity
capacity <- acclimation - basal
```

### Capacity vs. basal
```{r}
plot(capacity~basal)
summary(lm(capacity~basal))
cor(capacity,basal)
```
### Acclimation vs. basal
```{r}
plot(acclimation~basal)
summary(lm(acclimation~basal))
cor(acclimation,basal)
```

### Thoughts
Generally, we see that there is no relationship between survivorship in our acclimation treatment vs our basal treatment (which we would expect given the way we simulated the data). However, there was a very significant relationship betwenen acclimation capacity and the basal treatment. Specifically, we see a strong negative correlation that is being driven by the use of the basal tolerance variable in the calculation of the acclimation capacity (i.e. a spurious correlation driven by a variable being correlated with itself).
  Moreover, if we look at the R2 value we see that ~50% of the variation in our capacity variable can explained by the basal variable. This makes sense since we calculated acclimation capacity as an additive combination of the basal and acclimation variables, which were generated by drawing from identical normal distributions with equal variances. This means that on average we would only expect half of the variation to be due to the basal value in the calculation and half from the acclimation value. This also indicates that the strength of the spurious correlation could be driven by the relative amount of variance in the basal vs. acclimation components of our acclimation capacity calculation.

## Quick examination of spurious correlation strengths under a range of basal tolerance variances

Here we take a quick look at unequal variances among our basal vs. acclimation variables and its impact on the spurious correlation between acclimation capacity and basal tolerance:

We do this by simulating a basal tolerance variable that has half (0.5),equal (1), twice (2), triple (3), and five times the standard deviation as our acclimation variable (5).
```{r}
library(reshape,quietly = TRUE)
library(ggplot2,quietly = TRUE)
library(ggpmisc,quietly = TRUE)
library(cowplot,quietly = TRUE)

# This is to simulate variable levels of variance in our basal variable (while keeping the acclim
# ation value unchanged)
rel_deviation <-c(0.5,1,2,5)

bas_draw <- function(x){return(rnorm(100,0,c(x)))}
basal <-sapply(rel_deviation,bas_draw)
acclimation <- rnorm(100,0,1)
capacity <- acclimation - basal
#Covert in to data frames
acc_df <- data.frame(acclimation)
cap_df <- data.frame(capacity)
bas_df <- data.frame(basal)
# Scale 
bas_df <- data.frame(sapply(bas_df,scale))
cap_df <- data.frame(sapply(cap_df,scale))

## Convert into long form
# Add temp column names
colnames(bas_df) <- rel_deviation
colnames(cap_df) <- rel_deviation
# Melt
basal_long <- melt(bas_df)
cap_long <- melt(cap_df)
# Combine acclimation,basal,capacity value into single long dataframe
all_long <- cbind(acc_df,basal_long,cap_long[,2])
colnames(all_long) <- c("Acclimation_Value","Relative_Basal_Variance","Basal_Value","Capacity_Value")

my_eq <- y~x
ggplot(all_long,aes(y=Capacity_Value,x=Basal_Value,colour=as.factor(Relative_Basal_Variance))) +
  geom_smooth(method="lm",formula=my_eq) +
   stat_poly_eq(formula = my_eq, 
               aes(label = paste(..rr.label.., sep = "~~~")), 
               parse = TRUE) +
  labs(x = "Basal Value",y="Capacity Value",title="Basal vs Capacity correlation") +
  geom_point() + 
  facet_grid(.~as.factor(Relative_Basal_Variance)) +
  theme_cowplot() + 
  theme(legend.position = "none")

corr_vec <- NULL
for(i in 1:ncol(bas_df)){
  corr_vec <- c(corr_vec,cor(bas_df[,i],cap_df[,i]))
}
names(corr_vec) <- rel_deviation
```

### Estimated correlation between acclimation capacity and basal tolerance
```{r,echo=FALSE}
corr_vec
```

### Thoughts  
The strength of the correlation increases as the variance in the basal tolerance variable increases relative to the acclimation variable. This may be particularly problematic in the context of the Gerken paper if acclimation not only increases survivorship but also decreases the variance in survivorship among lineages. Without the data it is difficult to say with certainty if this is happening. However, looking at their barplot of survivorship to acute stress (S2.1) for each treatment (basal and short and long term acclimation) it does seem that the basal treatments tend to have equal or often times larger variances than the corresponding acclimation treatments. In particular, DACC has one of the lowest relative variances and DACC capacity also has one of the highest correlations with basal tolerance, which is consistent with this pattern being driven by a spurious correlation. 


